---
title: "BV-BRC Metadata Statistics"
author: "Abhirupa Ghosh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BV-BRC Metadata Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette summarizes the completeness of the bacterial metadata table served by BV-BRC. It pulls the full bacterial-genome metadata dump from BV-BRC via the project's CLI Docker image, normalizes character columns (UTF-8 fixes, whitespace trimming, mapping placeholder strings like `"NA"`, `"unknown"`, and `"not reported"` to actual `NA`s), and reports per-column missingness plus block-level presence/absence for host and geographic fields. The goal is to help decide which BV-BRC columns are worth featurizing downstream.

```{r setup, include=FALSE}
library(dplyr)
library(stringr)
library(purrr)
library(tibble)
```

## Helper functions

The chunk below defines all helpers used in the rest of the vignette:

- `fetchCompleteBVBRCMetadata()` — calls `docker run --rm danylmb/bvbrc:5.3 p3-all-genomes` twice: once to enumerate the available attributes, then again to pull the full bacterial-genome metadata table with those attributes selected. Returns a tibble with all character columns coerced to UTF-8.
- `is_placeholder_na()` and `clean_all_char_cols()` — normalize text columns by trimming whitespace, fixing UTF-8 glitches, and treating sentinel strings (`"NA"`, `"unknown"`, `"not reported"`, etc.) as `NA` rather than valid values.
- `summarize_block()` — computes per-row presence/absence summaries across a group of related columns (e.g. all host columns or all geographic columns).

```{r functions, echo=FALSE}
# function to download BVBRC metadata for all Bacteria
fetchCompleteBVBRCMetadata <- function(
  image = "danylmb/bvbrc:5.3",
  verbose = TRUE
) {
  if (isTRUE(verbose)) {
    message("Building attribute list from BV-BRC...")
  }

  # Step 1: get cleaned attribute list
  attr_cmd <- glue::glue(
    "docker run --rm {image} p3-all-genomes -f | ",
    "tail -n +2 | ",
    "sed -E 's/ \\((multi|related|derived)\\)//g' | ",
    "tr '\\n' ',' | sed 's/,$//'"
  )

  ATTRS <- tryCatch(
    system(attr_cmd, intern = TRUE),
    error = function(e) {
      stop("Failed to retrieve attribute list.\n", conditionMessage(e))
    }
  )

  if (length(ATTRS) == 0L) {
    stop("Attribute list is empty.")
  }

  if (isTRUE(verbose)) {
    message("Fetching BV-BRC bacterial genome metadata (this may take time)...")
  }

  # Step 2: use ATTRS in main query
  cmd <- glue::glue(
    "docker run --rm {image} p3-all-genomes ",
    "--in superkingdom,Bacteria ",
    "--attr \"{ATTRS}\""
  )

  raw_data <- tryCatch(
    system(cmd, intern = TRUE, ignore.stderr = TRUE),
    error = function(e) {
      stop("BV-BRC CLI call failed.\n", conditionMessage(e))
    }
  )

  if (length(raw_data) == 0L) {
    stop("BV-BRC returned no data. The service may be unavailable.")
  }

  # Step 3: parse safely
  df <- utils::read.table(
    text = raw_data,
    sep = "\t",
    header = TRUE,
    fill = TRUE,
    quote = "",
    check.names = FALSE,
    comment.char = "",
    colClasses = "character"
  )

  df <- tibble::as_tibble(df) |>
    dplyr::mutate(dplyr::across(
      dplyr::everything(),
      ~ iconv(.x, from = "", to = "UTF-8", sub = "")
    ))

  if (isTRUE(verbose)) {
    message(glue::glue("Retrieved {nrow(df)} rows x {ncol(df)} columns."))
  }

  return(df)
}

# Use your fix_utf8() if defined; otherwise a no-op fallback
if (!exists("fix_utf8")) {
  fix_utf8 <- function(x) x
}

# Tokens that should count as missing for text fields
missing_tokens <- c(
  "", " ", "NA", "N/A", "n/a", "na", "null", "Null", "NULL",
  "unknown", "Unknown", "UNKNOWN",
  "not provided", "Not provided", "NOT PROVIDED",
  "not reported", "Not reported", "NOT REPORTED",
  "unavailable", "Unspecified", "unspecified", "missing"
)

is_placeholder_na <- function(x) {
  if (!is.character(x)) {
    return(rep(FALSE, length(x)))
  }
  str_trim(x) %in% missing_tokens
}

# Clean ALL character/factor columns in the data frame
clean_all_char_cols <- function(df) {
  df |>
    mutate(across(
      where(~ is.character(.x) || is.factor(.x)),
      ~ {
        x <- .
        if (is.factor(x)) x <- as.character(x)
        if (is.character(x)) {
          x <- fix_utf8(x)
          x <- str_squish(x) # trim + collapse internal spaces
          x[is_placeholder_na(x)] <- NA # "", "N/A", "unknown", etc. -> NA
        }
        x
      }
    ))
}

# --- Block-level summaries ---------------------------------------------------
summarize_block <- function(df, cols, block_name) {
  if (length(cols) == 0) {
    return(tibble(
      block = block_name,
      n_total = nrow(df),
      n_rows_all_missing = nrow(df),
      pct_all_missing = 100,
      n_rows_any_present = 0,
      pct_any_present = 0
    ))
  }
  df |>
    transmute(
      .row_any_present = if_any(all_of(cols), ~ !is.na(.x)),
      .row_all_missing = if_all(all_of(cols), is.na)
    ) |>
    summarise(
      block = block_name,
      n_total = n(),
      n_rows_all_missing = sum(.row_all_missing),
      pct_all_missing = 100 * n_rows_all_missing / n_total,
      n_rows_any_present = sum(.row_any_present),
      pct_any_present = 100 * n_rows_any_present / n_total,
      .groups = "drop"
    )
}
```

## Compute statistics

Here we fetch the BV-BRC bacterial metadata table, clean it, and compute four artifacts that drive the rest of the vignette: `col_stats` (one row per column with totals, missing counts, and distinct non-missing values), `host_block_stats` and `geo_block_stats` (per-block presence summaries across host and geographic columns respectively), and `diag_table` (a focused per-column view of just the host + geo columns).

```{r stats}
# download the bvbrc table
bvbrc <- fetchCompleteBVBRCMetadata()

# Apply to your table
bvbrc_clean <- clean_all_char_cols(bvbrc)
# --- Per-column stats --------------------------------------------------------
col_stats <- tibble(column = colnames(bvbrc_clean)) |>
  mutate(
    n_total = nrow(bvbrc_clean),
    n_missing = map_int(column, ~ sum(is.na(bvbrc_clean[[.x]]))),
    n_present = n_total - n_missing,
    pct_missing = 100 * n_missing / n_total,
    n_distinct_non_missing = map_int(column, ~ n_distinct(bvbrc_clean[[.x]], na.rm = TRUE))
  ) |>
  arrange(desc(pct_missing), column)

# --- Define blocks -----------------------------------------------------------
host_cols <- c(
  "genome.host_common_name",
  "genome.host_group",
  "genome.host_name"
)

geo_cols <- c(
  "genome.isolation_source",
  "genome.geographic_location",
  "genome.isolation_country",
  "genome.state_province",
  "genome.city",
  "genome.county",
  "genome.latitude",
  "genome.longitude"
)

host_cols <- intersect(host_cols, colnames(bvbrc))
geo_cols <- intersect(geo_cols, colnames(bvbrc))

host_block_stats <- summarize_block(bvbrc_clean, host_cols, "host")
geo_block_stats <- summarize_block(bvbrc_clean, geo_cols, "geographic")

# --- Quick diagnostics: check specific columns you care about ----------------
diag_cols <- unique(c(host_cols, geo_cols))
diag_table <- tibble(column = diag_cols) |>
  mutate(
    n_total = nrow(bvbrc_clean),
    n_missing = map_int(column, ~ sum(is.na(bvbrc_clean[[.x]]))),
    n_present = n_total - n_missing,
    pct_missing = 100 * n_missing / n_total
  ) |>
  arrange(desc(pct_missing), column)
```

## Per-column completeness

The table below shows how complete each column in the BV-BRC bacterial metadata is — sorted with the most sparsely populated columns first. Columns with very high `pct_missing` are unlikely to be useful downstream and may be worth dropping from featurization.

```{r col_stats_output}
cat("\n--- Per-column stats (whole table) ---\n")
print(col_stats, n = Inf)
```

## Host block summary

Presence of the host-related columns (`host_common_name`, `host_group`, `host_name`) treated as a block. `n_rows_any_present` is the number of rows where at least one host field is populated; `n_rows_all_missing` is the number where every host field is empty.

```{r host_block_output}
cat("\n--- Host block stats ---\n")
print(host_block_stats)
```

## Geographic block summary

Same idea as the host block, but across the geographic columns (`isolation_source`, `geographic_location`, `isolation_country`, `state_province`, `city`, `county`, `latitude`, `longitude`).

```{r geo_block_output}
cat("\n--- Geographic block stats ---\n")
print(geo_block_stats)
```

## Diagnostics: per-column breakdown of host + geo

Column-by-column missingness for just the host and geographic fields, so you can see which specific columns drive the block-level numbers above.

```{r diag_output}
cat("\n--- Diagnostics for host + geo columns ---\n")
print(diag_table, n = Inf)
```
