---
title: "BV-BRC Metadata Statistics"
author: "Abhirupa Ghosh"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BV-BRC Metadata Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette summarizes the completeness of the bacterial metadata table served by BV-BRC. It uses the full bacterial-genome metadata table from the BV-BRC REST API (downloaded with the package's internal helpers and cached in the BiocFileCache), normalizes character columns (UTF-8 fixes, whitespace trimming, mapping placeholder strings like `"NA"`, `"unknown"`, and `"not reported"` to actual `NA`s), and reports per-column missingness plus block-level presence/absence for host and geographic fields. The goal is to help decide which BV-BRC columns are worth featurizing downstream.

## Helper functions

The download and cleaning helpers (`.fetchAllBVBRCmetadataApi()`, `.cleanBVBRCmetadata()` and `.loadBVBRCmetadata()`) are internal functions in `R/bvbrc_metadata.R`. They page through the BV-BRC genome API (`superkingdom = Bacteria`) and normalize text columns (UTF-8, whitespace, placeholder strings such as `"NA"`, `"unknown"`, `"not reported"` → `NA`). The cleaned table is cached as parquet in the BiocFileCache (`.loadBVBRCmetadata()`). The download takes a while, so the completeness sections below run only when that cache exists or the environment variable `AMRDATA_FETCH_BVBRC` is set to `true`. Only the block-summary helper is defined here:

- `summarize_block()` — computes per-row presence/absence summaries across a group of related columns (e.g. all host columns or all geographic columns).

```{r functions, echo=FALSE}
# The full BV-BRC table is large: use the BiocFileCache copy when present, and
# only download it on request:
#   Sys.setenv(AMRDATA_FETCH_BVBRC = "true")
have_cache <- tryCatch(
  !is.na(amRdata:::.bvbrcCachedPath()),
  error = function(e) FALSE
)
fetch_bvbrc <- identical(Sys.getenv("AMRDATA_FETCH_BVBRC"), "true")
knitr::opts_chunk$set(eval = have_cache || fetch_bvbrc)

# --- Block-level summaries ---------------------------------------------------
summarize_block <- function(df, cols, block_name) {
  if (length(cols) == 0) {
    return(tibble::tibble(
      block = block_name,
      n_total = nrow(df),
      n_rows_all_missing = nrow(df),
      pct_all_missing = 100,
      n_rows_any_present = 0,
      pct_any_present = 0
    ))
  }
  df |>
    dplyr::transmute(
      .row_any_present = dplyr::if_any(dplyr::all_of(cols), ~ !is.na(.x)),
      .row_all_missing = dplyr::if_all(dplyr::all_of(cols), is.na)
    ) |>
    dplyr::summarise(
      block = block_name,
      n_total = dplyr::n(),
      n_rows_all_missing = sum(.row_all_missing),
      pct_all_missing = 100 * n_rows_all_missing / n_total,
      n_rows_any_present = sum(.row_any_present),
      pct_any_present = 100 * n_rows_any_present / n_total,
      .groups = "drop"
    )
}
```

## Compute statistics

Here we download and clean the BV-BRC bacterial metadata table (from the BiocFileCache, or downloaded when `AMRDATA_FETCH_BVBRC=true`) and compute four artifacts that drive the rest of the vignette: `col_stats` (one row per column with totals, missing counts, and distinct non-missing values), `host_block_stats` and `geo_block_stats` (per-block presence summaries across host and geographic columns respectively), and `diag_table` (a focused per-column view of just the host + geo columns).

```{r stats}
bvbrc_clean <- amRdata:::.loadBVBRCmetadata()

# --- Per-column stats --------------------------------------------------------
col_stats <- tibble::tibble(column = colnames(bvbrc_clean)) |>
  dplyr::mutate(
    n_total = nrow(bvbrc_clean),
    n_missing = purrr::map_int(column, ~ sum(is.na(bvbrc_clean[[.x]]))),
    n_present = n_total - n_missing,
    pct_missing = 100 * n_missing / n_total,
    n_distinct_non_missing = purrr::map_int(column, ~ dplyr::n_distinct(bvbrc_clean[[.x]], na.rm = TRUE))
  ) |>
  dplyr::arrange(dplyr::desc(pct_missing), column)

# --- Define blocks -----------------------------------------------------------
host_cols <- c(
  "host_common_name",
  "host_group",
  "host_name"
)

geo_cols <- c(
  "isolation_source",
  "geographic_location",
  "isolation_country",
  "state_province",
  "city",
  "county",
  "latitude",
  "longitude"
)

host_cols <- intersect(host_cols, colnames(bvbrc_clean))
geo_cols <- intersect(geo_cols, colnames(bvbrc_clean))

host_block_stats <- summarize_block(bvbrc_clean, host_cols, "host")
geo_block_stats <- summarize_block(bvbrc_clean, geo_cols, "geographic")

# --- Quick diagnostics: check specific columns you care about ----------------
diag_cols <- unique(c(host_cols, geo_cols))
diag_table <- tibble::tibble(column = diag_cols) |>
  dplyr::mutate(
    n_total = nrow(bvbrc_clean),
    n_missing = purrr::map_int(column, ~ sum(is.na(bvbrc_clean[[.x]]))),
    n_present = n_total - n_missing,
    pct_missing = 100 * n_missing / n_total
  ) |>
  dplyr::arrange(dplyr::desc(pct_missing), column)
```

## Per-column completeness

The table below shows how complete each column in the BV-BRC bacterial metadata is — sorted with the most sparsely populated columns first. Columns with very high `pct_missing` are unlikely to be useful downstream and may be worth dropping from featurization.

```{r col_stats_output}
cat("\n--- Per-column stats (whole table) ---\n")
print(col_stats, n = Inf)
```

## Host block summary

Presence of the host-related columns (`host_common_name`, `host_group`, `host_name`) treated as a block. `n_rows_any_present` is the number of rows where at least one host field is populated; `n_rows_all_missing` is the number where every host field is empty.

```{r host_block_output}
cat("\n--- Host block stats ---\n")
print(host_block_stats)
```

## Geographic block summary

Same idea as the host block, but across the geographic columns (`isolation_source`, `geographic_location`, `isolation_country`, `state_province`, `city`, `county`, `latitude`, `longitude`).

```{r geo_block_output}
cat("\n--- Geographic block stats ---\n")
print(geo_block_stats)
```

## Diagnostics: per-column breakdown of host + geo

Column-by-column missingness for just the host and geographic fields, so you can see which specific columns drive the block-level numbers above.

```{r diag_output}
cat("\n--- Diagnostics for host + geo columns ---\n")
print(diag_table, n = Inf)
```

## Potential species for exploring the package

Candidate species were selected using the following criteria (built by `data_raw/species_count_summary.R` and shipped as the `species_count_summary` dataset):

- `antimicrobial_resistance_evidence` = "AMR Panel"
- `genome_quality` = "Good"
- `genome_status` = "WGS" or "Complete"
- `checkm_completeness` >= 98
- `genome_count` between 20 and 1,000 (inclusive)
- `antimicrobial_resistance` is not missing

```{r bvbrc-amr-species, eval=TRUE}
data("species_count_summary", package = "amRdata")

knitr::kable(
  species_count_summary,
  caption = "Recommended bacteria to include in Zenodo, with qualifying genome counts."
)
```
