---
title: "Introduction to amRdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to amRdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(amRdata)
```

## Overview

`amRdata` is part of a 3-package suite for predicting antimicrobial resistance (AMR) using machine learning models trained on bacterial genomic features.

This vignette demonstrates how to download, process and visualize metadata and genomics data. 

## Data Curation

For meta data curation with `amRdata`, use `retrieveMetadata()` to create a DuckDB containing metadata tables.

```{r}
# Download all AMR data for a species from BV-BRC
retrieveMetadata(user_bacs = "Shigella flexneri",
                             filter_type = "AMR",
                             base_dir = ".",
                             abx = "All",
                             overwrite = FALSE,
                             image = "danylmb/bvbrc:5.3",
                             verbose = TRUE)
```

This wrote tables 'amr_phenotype', 'genome_data', and 'metadata' to a DuckDB

For downloading genomes with proper AMR phenotype data, use `retrieveGenomes()`

```{r }
retrieveGenomes <- function(base_dir = ".",
                            user_bacs = "Shigella flexneri",
                            method = c("ftp", "cli"),
                            image = "danylmb/bvbrc:5.3",
                            skip_existing = TRUE,
                            ftp_workers = 8L,
                            cli_fasta_workers = 4L,
                            cli_gff_workers = 4L,
                            chunk_size = 50L,
                            verbose = TRUE)
```
This returns character vector of genome IDs and wrote complete file sets on disk.

to write a .txt file with downloaded genome file paths (.fna, .faa, .gff) 

```{r }
genomeList(base_dir = ".",
                       user_bacs = "Shigella flexneri",
                       verbose = TRUE)
```

#### A wrapper for downloading genomes and listing the paths
Internally runs:
1. `retrieveGenomes()`  
2. `genomeList()`  

```{r }
prepareGenomes(user_bacs = "Shigella flexneri",
                           base_dir = ".",
                           method = c("ftp", "cli"),
                           overwrite = FALSE,
                           verbose = TRUE)
```

## Data Processing
#### A wrapper for creating pangenome, protein clusters and annotate domains
Internally runs:
1. `.runPanaroo2Duckdb()` -> run Panaroo (optional panaroo-merge) to generate pangenome and create genomes by genes table
2. `CDHIT2duckdb()` -> run CDHIT to cluster proteins and create genomes by proteins table 
3. `domainFromIPR()` -> run IPRScan to annotate proteins to domains and create genomes by domains table
4. `cleanData()` -> Clean metadata and export the feature tables to Parquet + Parquet-backed DuckDB

```{r}
runDataProcessing(duckdb_path = "data/",
                              output_path = NULL,
                              # unified threads for all tools
                              threads = 16,
                              # Panaroo
                              panaroo_split_jobs = FALSE,
                              panaroo_core_threshold = 0.90,
                              panaroo_len_dif_percent = 0.95,
                              panaroo_cluster_threshold = 0.95,
                              panaroo_family_seq_identity = 0.5,
                              # CD-HIT
                              cdhit_identity = 0.9,
                              cdhit_word_length = 5,
                              cdhit_memory = 0,
                              cdhit_extra_args = c("-g","1"),
                              cdhit_output_prefix = "cdhit_out",
                              # InterPro
                              ipr_appl = c("Pfam"),
                              ipr_threads_unused = NULL,
                              ipr_version = "5.76-107.0",
                              ipr_dest_dir = "inst/extdata/interpro",
                              ipr_platform = "linux/amd64",
                              auto_prepare_data = TRUE,
                              # Metadata cleaning
                              ref_file_path = "data_raw/",
                              verbose = TRUE)
```

## Plots

Simple stats and plots to explore metadata

```{r }
generateSummary("data/metadata_parquet", 
                            out_path = "data/")
generatePlots("data/metadata_parquet", 
                            out_path = "data/")
```
