---
title: "Introduction to amRdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to amRdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(amRdata)
```

## Overview

`amRdata` is part of a 3-package suite for predicting antimicrobial resistance (AMR) using machine learning models trained on bacterial genomic features.

This vignette demonstrates how to download, process and visualize metadata and genomics data.

## Data Curation

For metadata curation with `amRdata`, use `retrieveMetadata()` to create a DuckDB containing metadata tables.

```{r}
# Download all AMR data for a species from BV-BRC
retrieveMetadata(
  user_bacs = "Shigella flexneri",
  filter_type = "AMR",
  base_dir = "../data/",
  abx = "All",
  overwrite = FALSE,
  image = "danylmb/bvbrc:5.3",
  verbose = FALSE
)
```

This wrote tables 'amr_phenotype', 'genome_data', and 'metadata' to a DuckDB

For downloading genomes with paired AMR phenotype data, use `retrieveGenomes()`

```{r }
retrieveGenomes(
  base_dir = "../data/",
  user_bacs = "Shigella flexneri",
  method = c("cli"),
  image = "danylmb/bvbrc:5.3",
  skip_existing = TRUE,
  ftp_workers = 8L,
  cli_fasta_workers = 4L,
  cli_gff_workers = 4L,
  chunk_size = 50L,
  verbose = TRUE
)
```

This returns character vector of genome IDs and wrote complete file sets on disk.

To write a .txt file listing downloaded genome filepaths (.fna, .faa, .gff)

```{r }
genomeList(
  base_dir = "../data/",
  user_bacs = "Shigella flexneri",
  verbose = TRUE
)
```

#### A wrapper for downloading genomes and listing the paths

Internally runs: 
1. `retrieveGenomes()`\
2. `genomeList()`

Allows users to input a species or taxon ID and automate all data downloading and curation steps.

```{r }
prepareGenomes(
  user_bacs = "Shigella flexneri",
  base_dir = "../data/",
  method = c("cli"),
  overwrite = FALSE,
  verbose = TRUE
)
```

## Data Processing

#### A wrapper for creating pangenome, protein clusters and annotate domains

Internally runs: 
1. `runPanaroo2Duckdb()` -\> run Panaroo (optional panaroo-merge) to generate pangenome and create genomes by genes table 
2. `CDHIT2duckdb()` -\> run CDHIT to cluster proteins and create genomes by proteins table 
3. `domainFromIPR()` -\> run IPRScan to annotate proteins to domains and create genomes by domains table 
4. `cleanData()` -\> Clean metadata and export the feature tables to Parquet + Parquet-backed DuckDB

```{r}
runDataProcessing(
  duckdb_path = "../data/Shigella_flexneri/Sfl.duckdb",
  output_path = NULL,
  # unified threads for all tools
  threads = 16,
  # Panaroo
  panaroo_split_jobs = FALSE,
  panaroo_core_threshold = 0.90,
  panaroo_len_dif_percent = 0.95,
  panaroo_cluster_threshold = 0.95,
  panaroo_family_seq_identity = 0.5,
  # CD-HIT
  cdhit_identity = 0.9,
  cdhit_word_length = 5,
  cdhit_memory = 0,
  cdhit_extra_args = c("-g", "1"),
  cdhit_output_prefix = "cdhit_out",
  # InterPro
  ipr_appl = c("Pfam"),
  ipr_threads_unused = NULL,
  ipr_version = "5.76-107.0",
  ipr_dest_dir = "inst/extdata/interpro",
  ipr_platform = "linux/amd64",
  auto_prepare_data = TRUE,
  # Metadata cleaning
  ref_file_path = "../data_raw/",
  verbose = TRUE
)
```

## Plots

Simple stats and plots to explore metadata

```{r }
generateSummary("data/metadata_parquet",
  out_path = "data/"
)
generatePlots("data/metadata_parquet",
  out_path = "data/"
)
```
